How to select predictive models for decision-making or causal inference?

### 2024-10-20 response_to_reviewer.md

*or some other that can be easily found in literature.*

> We thank the reviewer for the references. We have added several
> references for Causal ML into the manuscript, in particular [Kaddour et al., 2022](https://arxiv.org/pdf/2206.15475), corresponding to
> https://www.causal-machine-learning.com/, and [Chernozhukov et al., 2024](https://causalml-book.org/) which both cover most of the
> subfields at the intersection of causal inference and machine learning.

# Reviewer 2

*Overall assessment:*

*Let me begin with positive comments.  I think the paper is very important because it describes a sophisticated methodology for experimental examination of causal models.
Despite some attempts this is still lacking in literature.  I especially like the simulation scheme which uses different features for data generation and learning models.  Moreover thorough experimental evaluations of causal models are not very frequent making the paper important.  Code is available on Github which is a big plus.*

*I am however not sure about the overall conclusions, which are that R-risk is always the best.  Performance of risk functions may depend on several factors and the overall best performance might be a result of bias in available data.  The only factor taken into account by the authors is overlap.  However, only one type of lack of overlap is tested in experiments, based on shifted Gaussians.  This will likely lead to violation of the positivity assumption.  How about different scenarios, e.g.: perfect overlap but different treatment/control distributions?  Confounding need not necessarily lead to lack of overlap.  Another possible factor is the ratio of causal effect to background responses.  It is known that it may affect relative performance of HTE estimators.*

*Overall: I am no fully convinced that R-risk is always the best.  The answer could be dependent on other factors and I hope the authors would explore more scenarios.*

> We thank the reviewer for the thoughful comments and the call for
> nuance. Our first response is that our thoughts did not come out right
> and that it is not clear, indeed, that R-risk is always the best.
> Rather, that it is a good default solution. We have rewritten parts of
> the paper to convey this nuance better.

> In addition, the reviewer is correct that factors beyond overlap
> probably affect which risk is best. We have followed the suggestion to
> explore more diverse simulation scenarii, as detailing the sensitivity
> of our result on the R-risk makes a more nuanced and stronger analysis.
> Thus, we ran a dedicated experiment focusing on the ratio of causal
> effect to background response. We measure the ratio by the absolute
> mean difference between the causal effect and the baseline response,
> measured empirically on each dataset instance $\Delta_{\mu} =
> \frac{1}{N} \sum_{i=1}^N \big | \frac{\mu_{1}(x_i) -
> \mu_{0}(x_i)}{\mu_{0}(x_i)} \big|$. We vary a dedicated parameter
> ($\omega$ in Algorithm 2) in the simulation to force this ratio to
> cover a wide range of values. In 90 dataset instance, it ranges several
> orders of magnitude from 0.005 to 9675. Figure 7 gives the main results
> of this experiment. We see that R-risk is still the best risk on
> average, but when the effect ratio is high (tertile between 9 and
> 9675), the reweighted mu-risk and the the tau-risk_IPW are better than
> the R-risk. However, for low value of the ratio,  the R-risk largely
> outperforms all other metrics. We added these new results to the main
> paper. We also nuanced our conclusion on the R-risk, underlying that it
> is better on average but that in some situations, other risks might be
> more adapted. A typical such situation would be when the applied
> scientist is interested in estimating an effect with a big ratio and
> want to focus on the part of the population where there is low overlap.

@@

> We did not explore more sophisticated treatment/control distributions for our simulation. Building these new data generation processes would imply a big modification to our setup. We rely on the easy inversion of the distribution functions of the gaussians (p(X|A=a)) to analytically compute the true propensity score. In addition, we think that the semi-simulated datasets already provide some interesting argument in favor of r-risk with more more diverse treatment/control covariate distributions.

*Another question: do the differences between risks in experiments follow known theoretical results?  For example, it is well known (e.g. https://www.semanticscholar.org/reader/112aba9c62f98b9684693f4c20c30274bc9eb1dc or [Rudas ans jaroszewisz, 2017](https://link.springer.com/content/pdf/10.1007/s10618-018-0576-8.pdf)) that the tau risk performs poorly.*

@@

> The second paper is from the uplift regression literature. It focus on full randomization problems and a linear specification with gaussian predictors. It shows theoretically that a linear T-learner (called double model approach in their approach) has smaller asymptotical variance than a treatment ratio reweighted outcome S learner. However, they note that in non normal cases the  T-learner can be ill-defined or with high variance. They propose a corrected reweighted S-learner approach to reduce this  variance. They show empirically that it behaves better than the non corrected version on a simulation and two real world problems. 
> I am not sure to see the link with the tau-risk. The paper does not compare risks but corrected or uncorrected vs T-learners. In addition, they only consider full randomization and linear problems. In our case, we consider a more general setting with a variety of data generation procedures and we focus on non-randomized situations where the distribution of the treatment is conditioned by the predictors.

*Another problem is that the paper is hard to follow. The authors rapidly switch between topics: risk, selection, nuisance parameter estimation so the reader quickly gets lost.  The sections in supplementary materials contain important information but not all are referenced in main text.  The paper gives an impression of several experiments put together in some order without guiding the reader through the maze.  I think organization should be improved before publication.*

@@

> We tried to better guide the reader through the method section and to clarify the motivation behind the different experiments. 
> We checked the supplementary materials but we did not find any appendix that was not referenced at least once in the main text (several are referred in figure captions). Did you have a specific appendix in mind that would require a reference in the main text?

Other comments:
---------------

*In the Methods section g-computation is described.  It seems indeed the first name under which the technique appeared in literature, but other names, most importantly S-learner are being used frequently nowadays.  Interestingly those names are mentioned in Empirical study on page 4, however, the close relationship (equivalence?) between S-learner and g-computation is not mentioned.*

@@

> Thanks for the remark. We corrected a typo in the equation. There should not have any mention of X in the conditional expectation.

*page 17: proposition number is missing*

> Thanks for this remark. We fixed it.

*A.7: the data generation procedure needs better explanations.  Why not call (7) P(d|A) since that's how its used later?  On page 17 you say '[b1,bD]~P(x)' and later 'x~P(x)' which causes confusion.  Also, how are beta vectors generated?*

@@

> We assumed that you talked about the equation 17 rather than 7. Could you point where we use the P(d|A) notation? We did not find it in the manuscript. 
> The notation b1, ..bD ~P(x) was meant to be understood as the beta vectors are drawn from a distribution P(x). The later mention of 'x~P(x)' was meant to be understood as the definition if the RBF function. We removed the second mention which causes confusion.
> The beta vectors are drawn from two normal distributions with mean 0 and unit variances. We added this information in the manuscript.
> We added the Algorithm 2, hopping that it would help the reader to understand exactly how we generate the data for our simulation.

*page 4: TLearner uses different feature representation for each treatment arm.  Does this really happen in practice?  Usually all subgroups are described with the same set of features.*

@@

> It is indeed true that in usual practice all subgroups are described with the same set of features. However, as soon as we are using a machine learning (necessary to model CATE) such as a random forest or boosted trees, the inner representations of the features inside the model is different for both population. This idea has been exploited to build representation aware neural net models for cate estimation by [Johansson et al., 2022](https://www.jmlr.org/papers/volume23/19-511/19-511.pdf). It is also the case for simpler models such as a linear model with treatment interaction and lasso regularization.

*page 4/5: Family of candidate estimators: why are only linear models used in some dataset and forests for other?  The leaf sizes of forests are chosen from a rather small range, why not use some percentage of the size of the training set?*

@@

> For the simulations we chose estimators that are out of the functional class of the  true data generation process (a candidate estimator cannot model exactly the data generation process) but that are quite similar to it. They are out of the functional class because the representers are different from the one of the true data (they are two randomly sampled data points). They are close because they have the same functional form : a linear model on top of two data representers. We used the simulations to explore the wider range of data generation processes, focusing on a gradient of overlap. Because of the sheer number of generated dataset, we did not include computationally costly boosted trees in the candidate family.

> The choice of the candidate family of estimators was guided by the idea to compare estimators that are not too different from each other, making the model selection task more challenging. The important point is that we are not focusing on the best possible estimator but on the best possible model selection procedure. To identify the optimal procedure, it is more convincing to compare close candidate estimators than a wide variety of candidate estimators. In the later case, one estimator could dominate all other estimators for every selection procedure. This explains why we are choosing a narrow range of leaf sizes for the forests.

Language comments:
------------------

*It seems that English might need improvement in some cases, for
example:*

*Abstract: "We study such a variety of model selection procedures in practical settings": a "large variety" seems more appropriate.*

*On page 1, the sentence "Predictive modeling bridges to causal inference theory under the name of outcome models ..." contains "bridges under the name" which also sounds strange.*

*On page 2: In the sentence "Then we anchor causal model selection in the potential outcome framework and details the causal risks and model-selection procedure." I believe 'details' should be 'detail'.*

*Algorithm 1: 'a candidate estimators f' should be 'a candidate estimator f'*

*throughout: "an RBF expansion" is more commonly used than "a RBF expansion"*

*page 4: 'for both population' should be 'for both populations'*

### 2024-10-20 how_to_select_latex.tex
iter

The simulations have been designed to evaluate the effect of the overlap parameter. The semi-simulated datasets have been included to explore more diverse and noisy covariate distributions. These simulations cover a wide range of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neural network, or language
models.

\paragraph{Limitations}

On average, the R-risk outperforms other risks, but it is not optimal for every
situations. If one possesses some prior knowledge on the data generation process,
another risk might be more adapted. This is typically the case for big causal
effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
should be better.

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds.pdf}
    %\hfill
    \caption{\textbf{The $R$-risk is the best metric}: Relative Kendall's $\tau$
        agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to
        the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}.
    }\label{apd:fig:relative_kendalls_all_datasets_all_metrics}
\end{figure}

>>>>

The simulations, designed to evaluate the effect of the overlap
parameter, also explore more diverse and noisy covariate distributions.
They cover a diversity of causal settings such as different ratio of causal effect to background responses, and functional links between covariates, outcome and treatment.

@@

Increasingly complex prediction models integrating richer medical data
have flourished because their predictions can be easily
demonstrated and validated on left-out data. But using them to underpin a
decision on whether to treat or not requires more careful validation,
using a metric accounting for the
putative intervention, the $R\text{-risk}$. On average, the $R\text{-risk}$
brings a sizeable benefit to select the most adequate model, even when
model development is based on treated and
untreated population with little differences, as in RCTs. 
%
Our conclusions are that without prior knowledge, the R-risk is a good
default. However, there is much remaining variation, and the R-risk will
not be optimal for every situation. We have identified one such specific
situation: when the causal effect is large compared to the variation of
the baseline effect.

%If one possesses some prior knowledge on the data generation process,
%another risk might be more adapted. This is typically the case for big causal
%effect ratio to baseline and low overlap where the $\tau\text{-risk}_{IPW}$
%should be better.

To facilitate better model selection, we provide Python
code\footnote{\url{https://github.com/soda-inria/causal_model_selection}}.
% Using the $R\text{-risk}$ does make evaluation more
% complicated not only because the procedure is more involved, but also
% because each intervention requires a dedicated evaluation. However, such
% off-policy evaluation remains much less costly than the recommended good
% practice of impact evaluation testing the ability of a prediction model
% to actually guide patient health \cite{hendriksen2013diagnostic}. Also,
This model-selection procedure puts no constraints on the models used to
build predictive models: it opens the door to evaluating a wide range of
models, from gradient boosting to convolutional neural network, or language
models.

@@

\begin{figure}[!b]
    \centering
    \includegraphics[width=\linewidth]{_1_r_risk_domination_mrisks_by_ds.pdf}
    %\hfill
    \caption{\textbf{The $R$-risk is the best metric on average}: Relative Kendall's $\tau$
        agreement with $\tau\text{-risk}$. Strong and Weak overlap correspond to
        the first and last tertiles of the overlap distribution measured with
        Normalized Total Variation eq. \ref{eq:ntv}.
    }\label{apd:fig:relative_kendalls_all_datasets_all_metrics}
\end{figure}
